Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-19 projectReport/abstract.tex
All G edits done

    In this project, the base assumption is that data is originates from  some discrete-time dynamical system, particularly one of a chaotic nature, but no assumptions are made pertaining to a specific dynamical system.
    A general learning problem is formulated and a well-known approach to the problem, alongside its practical limitations, is then considered.
    The notion of driven dynamical systems is introduced, and we subsequently establish the theoretical existence of a dynamical system possessing certain properties which guarantees it to be topologically conjugate (`dynamically equivalent') to the dynamics of the underlying system from which measurements were originally taken. 
    Our methodology is implemented practically by making use of recurrent neural networks to learn a map topologically equivalent to the underlying system that forecasts future states.
    In so doing we present the workability of an approach to obtain long-term topological and statistically consistent predictions for simple physical systems as well as a some chaotic attractors.
    \end{abstract}

>>>>

    In this project, the base assumption is that data originates from  some discrete-time dynamical system, particularly one of a chaotic nature, 
    but no assumptions are made pertaining to a specific dynamical system.
    A general learning problem is formulated and a well-known approach to the problem, alongside its practical limitations, is then considered.
    The notion of driven dynamical systems is discussed, and we establish the theoretical conditions that guarantee its single-delay dynamics to be topologically conjugate (`dynamically equivalent') to the dynamics of the underlying system from which measurements were originally taken. 
    The methodology is implemented practically by making use of recurrent neural networks and deep learning methods to learn a map that in turn helps forecast the future observations of the underlying system.
    In so doing we present the workability of an approach to obtain long-term topological and statistically consistent predictions for some chaotic attractors, and notably the double pendulum -- a system difficult to model from data.
    \end{abstract}

### 2022-01-19 projectReport/chapter2.tex
All G edits done

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. \ednote{Could you refer me to a resource here?}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to~\cite{saber2010introduction}.

@@

To this end, we make 3 remarks.
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
  In Graph~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted agains time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition}
  [\bf {Devaney's Chaos}]\label{Dfn_ChaosDec}\rm
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties.
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

@@

    \begin{figure}[ht]
      \includegraphics[scale=0.6]{Graphs/_tentmap_2.eps}
          \centering
          \captionof{figure}{Graph of the $T^2$ map with bases at $x=0$, $x=\frac{1}{2}$ and $x=1$ respectively.}\label{fig:T2tentmap}
    \end{figure}

@@

To show topological similarity (or sameness) between two metric or topological spaces, one needs to establish a homeomorphism between these two spaces. 
However, in the study of dynamical systems, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.
For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
Therefore, to find dynamically similar systems, one must establish a dynamical equivalence as illustrated in the simple commutativity diagram below~\ref{eqn_conjugacy}.

@@

\begin{Definition}
  [\bf {Conjugacy and Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
   Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$, where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

% \ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

>>>>

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed~\cite{saber2010introduction}. 
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to~\cite{saber2010introduction}.

@@

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
  In the plot Fig.~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted agains time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition}
  [\bf {Devaney's Chaos}]\label{Dfn_ChaosDec}\rm
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties:
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

@@

    \begin{figure}[ht]
      \includegraphics[width=0.5\linewidth]{Graphs/_tentmap_2.eps}
          \centering
          \captionof{figure}{Graph of the $T^2$ map with bases at $x=0$, $x=\frac{1}{2}$ and $x=1$ respectively.}\label{fig:T2tentmap}
    \end{figure}

@@

To show topological similarity (or sameness) between two metric or topological spaces, one needs to establish a homeomorphism between these two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.
For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
Therefore, to find dynamically similar systems, one must establish a dynamical equivalence as illustrated in the simple commutativity diagram below~\ref{eqn_conjugacy}.

@@

\begin{Definition}
  [\bf {Conjugacy and Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
  Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$ where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

### 2022-01-19 projectReport/chapter3.tex
All G edits done

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems that have unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem, and discuss various issues faced while forecasting.

@@

As already mentioned, the actual state $u$ of a system is nearly never fully known. The most one can do is to place probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem. 
\vspace{-6mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate; the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

@@

By generic, we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \ednote{B: Where can I find a citation for this?}. 
In our scenario -- and this remains important throughout -- the input space $U$ is considered the attractor.~\label{attractor_U}

@@

Though not uniquely related to the Takens embedding theorem, we opt to make mention of another requirement to our work here. 
The methods described in this project are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, there exists some point in $U - A$ that does not have a preimage such that $T$ cannot be not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the dying oscillations for a damped pendulum (see Figure~\ref{fig:damped_pendulum}). The double pendulum is discussed in greater depth  is provided in Chapter~\ref{ch5}.
The requirement is not overly restrictive as map that is topologically transitive on a subspace must also be surjective on it; if it were to be not surjective there would be an open set $V$ that does not intersect with any future iterates of an open set  $U$ disjoint from it.

>>>>

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems with unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem and discuss various issues faced while forecasting.

@@

As already mentioned, the actual state $u$ of a system is nearly never fully known. The most one can do is to place probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem:
\vspace{-6mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate; the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from \cite{takens1981detecting} after establishing preliminary definitions.

\begin{Definition}
  [\bf{Manifold, Compact Manifold}]\label{Dfn_compactManif}\rm
  A topological space is called a manifold if every point in the space has a neighbourhood homeomorphic to $\mathbb{R}^n$. Succinctly, a manifold is a locally Euclidean topological space.
  If a manifold is compact as a space topologically, then it is termed a compact manifold.
\end{Definition}

@@

By generic we mean a residual set on a certain topology on appropriate spaces of functions~\cite{takens1981detecting}.
% In our scenario -- and this remains important throughout -- the input space $U$ is considered the attractor.~\label{attractor_U}

@@

Though not uniquely related to the Takens embedding theorem, we opt to make mention of another requirement to our work here. 
The methods described in this project are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, there exists some point in $U - A$ that does not have a preimage such that $T$ cannot be not surjective. 
An example of this would be the case when a system displays energy-loss. 
Consider, for instance, consider the dying oscillations for a damped pendulum (see Figure~\ref{fig:damped_pendulum}). The double pendulum is discussed in greater depth  is provided in Chapter~\ref{ch5}.

The requirement is not overly restrictive as map that is topologically transitive on a subspace must also be surjective on it; if it were to be not surjective there would be an open set $V$ that does not intersect with any future iterates of an open set  $U$ disjoint from it.

### 2022-01-19 projectReport/chapter4.tex
All G edits done

\ednote{Moved to after defn of USP/ESP: The reachable set itself can defined independently from $g$ possessing the ESP/USP. } 
Note that $x\in{X_n(\overline{u})}$ if and only $g$ has a solution $\{x_k\}$ for $x_n=x$ and input $\overline{u}$, a result proved in nonautonomous dynamical systems literature~\cite{manjunath2014dynamics, manjunath2013echo}.

@@

\begin{Theorem}\label{Thm_CompactExistence}
 Let $g$ be a driven system.  If $X$ is compact then for each input $\overline{u}$, there exists at least one solution to the driven system $g(\ldotp, x)$
\end{Theorem}
\vspace{-6mm}
{\bf Proof.} May be found in~\cite{kloeden2011nonautonomous, manjunath2014dynamics, manjunath2013echo}
% One may easily construct many systems with trivial solution-sets, such as $g(u,x)=x$ which has only the constant solution $x$ and so for $U=[-1,1]$, the system would have no solution if $|x|>1$. To refine the scenario, we consider only systems with unique solutions.

@@

Secondly, it is worth noting that the reachable set~\ref{Dfn_ReachableSet} can be defined independently from $g$ possessing the USP. \ednote{hmm. does it fit here}
% Paraphrased.
%Additionally, $g$ having the USP is in our context also equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively also that $g$ exhibits the Echo State Property (ESP).

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to its surjectivity), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. (Compare this with Definition~\ref{defn_causallyEmbed}.) 
%\ednote{B: edited the sentence here}

@@

So far it is not clear whether the USP alone ensures the ability to causally embedding a dynamical system and we mention that a driving function $g$ cannot just be chosen in a frivolous manner. This could possibly complicate our task. %\ednote{B: Grammarly}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.35]{Graphs/_Nonlinearity_Fig_response.eps}
  \centering
\captionof{figure}{Input sequence (black) and a coordinate of a simulated solution $(x_0,x_1,\ldots,x_{5000})$ of a RNN  plotted in red (with parameter $a=1$, $\alpha=0.99$ where $g$ has the USP) and blue ($a=1$, $\alpha=1.05$ where $g$ does not have the USP) for randomly generated matrices $A$ and $B$. $B$ has unit spectral radius.  Reproduced from~\cite{Manju_Nonlinearity}. } 
\label{fig:memloss_conttime} 
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{Graphs/_summarypictorial.eps}
  \centering
  \captionof{figure}{Pictorial summary of results obtained up to now and how they overlap. Reproduced from~\cite{SIG25May} }
\label{fig:pictorialSummary}
\end{figure}

@@

To circumvent this: if a driven system with a saturation function \footnote{A function $f(t)$ that is monotonic and approaches a constant as $|t|\to \infty$, for e.g., $\tanh(t)$ on $\mathbb{R}$} is used, we can ensure the existence of an attractor $A$ of the map that implements 
$(\pi_2,g) \circ (\Gamma,\pi_2)$ (printed in the red on Figure~\ref{Scomm_imp})
 which in practice turns out to be defined on $Y_T^+$ has an attractor $A$  containing $Y_T$.
By restricting the dynamics to not spill beyond $A$, prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed here, may be found in~\cite{Supp}.
\begin{figure}[ht]
  \includegraphics[scale=0.25]{Graphs/_autovsdriven.eps}
  \centering
  \captionof{figure}{Illustrating empirically the importance of global dissipativity. Plotted below are the three principal components of randomly initiated points (blue) of a learnt version of $G_T$ versus the trajectory $(x_{n-1}, x_n)$ (red) of driven states obtained from the Fractal Dream Attractor's data~\ref{Thomas_Attractor}. }
 \label{fig:learningFailure}
\end{figure}

>>>>

Note that $x\in{X_n(\overline{u})}$ if and only $g$ has a solution $\{x_k\}$ for $x_n=x$ and input $\overline{u}$, a result proved in nonautonomous dynamical systems literature~\cite{manjunath2014dynamics, manjunath2013echo}.

@@

\begin{Theorem}\label{Thm_CompactExistence}
 Let $g$ be a driven system.  If $X$ is compact then for each input $\overline{u}$, there exists at least one solution to the driven system $g(\ldotp, x)$
\end{Theorem}
\vspace{-6mm}
{\bf Proof.} May be found in~\cite{kloeden2011nonautonomous, manjunath2014dynamics, manjunath2013echo}

@@

% Secondly, it is worth noting that the reachable set~\ref{Dfn_ReachableSet} can be defined independently from $g$ possessing the USP. \ednote{hmm. does it fit here}
% Paraphrased.
%Additionally, $g$ having the USP is in our context also equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively also that $g$ exhibits the Echo State Property (ESP).

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to its surjectivity), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. (Compare this with Definition~\ref{defn_causallyEmbed}.)

@@

So far it is not clear whether the USP alone ensures the ability to causally embed a dynamical system.
We also mention that a driving function $g$ cannot just be chosen in a frivolous manner. 
This could possibly complicate our task. %\ednote{B: Grammarly}

@@

\begin{figure}[ht]
  \includegraphics[width=0.5\linewidth]{Graphs/_Nonlinearity_Fig_response.eps}
  \centering
\captionof{figure}{Input sequence (black) and a coordinate of a simulated solution $(x_0,x_1,\ldots,x_{5000})$ of a RNN  plotted in red (with parameter $a=1$, $\alpha=0.99$ where $g$ has the USP) and blue ($a=1$, $\alpha=1.05$ where $g$ does not have the USP) for randomly generated matrices $A$ and $B$. $B$ has unit spectral radius.  Reproduced from~\cite{Manju_Nonlinearity}. } 
\label{fig:memloss_conttime} 
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=0.7\linewidth]{Graphs/_summarypictorial.eps}
  \centering
  \captionof{figure}{Pictorial summary of results obtained up to now and how they overlap. Reproduced from~\cite{SIG25May} }
\label{fig:pictorialSummary}
\end{figure}

@@

To circumvent this: if a driven system with a saturation function \footnote{A function $f(t)$ that is monotonic and approaches a constant as $|t|\to \infty$, for e.g., $\tanh(t)$ on $\mathbb{R}$} is used, we can ensure the existence of an attractor $A$ of the map that implements 
$(\pi_2,g) \circ (\Gamma,\pi_2)$ (printed in the red on Equation~\ref{Scomm_imp})
 which in practice turns out to be defined on $Y_T^+$ has an attractor $A$  containing $Y_T$.
By restricting the dynamics to not spill beyond $A$, prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed here, may be found in~\cite{Supp}.
\begin{figure}[ht]
  \includegraphics[scale=0.25]{Graphs/_autovsdriven.eps}
  \centering
  \captionof{figure}{Illustrating empirically the importance of global dissipativity. Plotted below are the three principal components of randomly initiated points (blue) of a learnt version of $G_T$ versus the trajectory $(x_{n-1}, x_n)$ (red) of driven states obtained from the Fractal Dream Attractor's data~\ref{Thomas_Attractor}. }
 \label{fig:learningFailure}
\end{figure}

### 2022-01-19 projectReport/chapter5.tex
All G edits done

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\ref{eqns_from_data}\ednote{How do I get both?}. To reach this point though, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value, lies within the range $[-1,1]$, and has been zero-padded as required (see paragraph below for further discussion), we drive the system with external input through the function $g$.

@@

We now turn our attention to another practical issue related to the forecasting of chaotic dynamical systems.  
Recall that the twin equations~\ref{eqns_from_data} generate a model from data. 
For a chaotic system, a small amount of noise due either to input-noise or computational error would cause solutions from even a very accurate model with the same initial conditions to diverge. %\ednote{Grammarly}
Statistically, however,  a chaotic system will behave well in the sense that the visiting frequency (the average time spent by an orbit in any region of the phase space of the dynamical system) is identical for almost all initial conditions. 
The phrase `almost all' here is specified as a measure and refers to a result in ergodic theory (\cite{wikiErgodicTheory}) which is a topic beyond the scope of this project.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modelling the data, we plot the densities of the trajectories of the actual and predicted solutions in ~\ref{fig:dp_success_density},~\ref{fig:noisydp_success_density} and ~\ref{fig:Thomas_TrajDensity}.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.95\linewidth]{Graphs/_dp_success_4coords_traj.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close up until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{Graphs/_Thomas_1.eps}\caption*{Predicted trajectories of the $x$- and $y$-coordinates for the Thomas attractor with parameter-value $\beta=0.1056$ demonstrate empirically the ability to predict the evolution of the trajectory for the next an estimated 100 timesteps into the future near-exactly. Here no noise was added.}
  \includegraphics[scale=0.5]{Graphs/_Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate ($x$-coordinate) of the dataset for the Cyclically Symmetric Attractor with parameter-value $\beta=0.1056$.}
\end{figure} \label{fig:Thomas_TrajDensity}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth,left]{Graphs/_Clifford_1_nonoise.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 1000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we are able to lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Graphs/_Clifford_2_nonoise.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Graphs/_Clifford_3_nonoise.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor (red) and the predicted values (blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{fig:Clifford}
\end{figure}

>>>>

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\eqref{eqns_from_data}\ednote{How do I get both?}. To reach this point though, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value, lies within the range $[-1,1]$, and has been zero-padded as required (see paragraph below for further discussion), we drive the system with external input through the function $g$.

@@

We now turn our attention to another practical issue related to the forecasting of chaotic dynamical systems.  
Recall that the twin equations~\ref{eqns_from_data} generate a model from data. 
For a chaotic system, a small amount of noise from even a very accurate model would cause solutions with the same initial conditions to diverge. This noise could be due either to input-noise or computational error.
Statistically, however,  a chaotic system will behave well in the sense that the visiting frequency (the average time spent by an orbit in any region of the phase space of the dynamical system) is identical for almost all initial conditions. 
The phrase `almost all' here is specified as a measure and refers to a result in ergodic theory (\cite{wikiErgodicTheory}) which is a topic beyond the scope of this project.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modelling the data, we plot the densities of the trajectories of the actual and predicted solutions in ~\ref{fig:dp_success_density},~\ref{fig:noisydp_success_density} and~\ref{fig:Thomas_TrajDensity}.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_traj.eps} 
  \captionof{figure}{Predicted (red) and actual (blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads. From top to bottom are as follows: $x$-coordinate of topmost pendulum head, $y$-coordinate of topmost pendulum head, $x$-coordinate of lower pendulum head, $y$-coordinate of lower pendulum head. 
  These graphs were constructed by predicting the DP 1000 timesteps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. Here we lock onto the near-exact trajectory for about 300 timesteps and stay close up until about 600 seconds.  } 
  \label{fig:dp_success_traj}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{Graphs/_dp_success_4coords_hist.eps}
  \captionof{figure}{Densities of predicted and actual values of $x$- and $y$-coordinate of pendulum heads.}
  \label{fig:dp_success_density}
 \end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{Graphs/_Thomas_1.eps}\caption*{Predicted trajectories of the $x$- and $y$-coordinates for the Thomas attractor with parameter-value $\beta=0.1056$ demonstrate empirically the ability to predict the evolution of the trajectory for the next an estimated 100 timesteps into the future near-exactly. Here no noise was added.}
  \includegraphics[scale=0.5]{Graphs/_Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate ($x$-coordinate) of the dataset for the Cyclically Symmetric Attractor with parameter-value $\beta=0.1056$.}
\end{figure}\label{fig:Thomas_TrajDensity}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,left]{Graphs/_Clifford_1_nonoise.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 1000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we are able to lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Graphs/_Clifford_2_nonoise.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Graphs/_Clifford_3_nonoise.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor (red) and the predicted values (blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{fig:Clifford}
\end{figure}
